---
title: "S. cerevisiae sample estimation, merged sample edition (20180606)."
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
 html_document:
  code_download: true
  code_folding: show
  fig_caption: true
  fig_height: 7
  fig_width: 7
  highlight: default
  keep_md: false
  mode: selfcontained
  number_sections: true
  self_contained: true
  theme: readable
  toc: true
  toc_float:
    collapsed: false
    smooth_scroll: false
---

<style>
  body .main-container {
    max-width: 1600px;
  }
</style>

```{r options, include=FALSE}
if (!isTRUE(get0("skip_load"))) {
  library("hpgltools")
  tt <- devtools::load_all("~/hpgltools")
  knitr::opts_knit$set(progress=TRUE,
                       verbose=TRUE,
                       width=90,
                       echo=TRUE)
  knitr::opts_chunk$set(error=TRUE,
                        fig.width=8,
                        fig.height=8,
                        dpi=96)
  old_options <- options(digits=4,
                         stringsAsFactors=FALSE,
                         knitr.duplicate.label="allow")
  ver <- "20180606"
  previous_file <- paste0("01_annotation-v", ver, ".Rmd")

  tmp <- try(sm(loadme(filename=paste0(gsub(pattern="\\.Rmd", replace="", x=previous_file), "-v", ver, ".rda.xz"))))
  rmd_file <- paste0("02_sample_estimation_merged-v", ver, ".Rmd")
}
```

S. cerevisiae sample estimation of all samples (old and new)
============================================================

This document should make clear the suitability of our yeast data for differential expression
analyses.  It should also give some ideas about the depth and distribution of
the data.

# Gathering samples

Later, we will likely choose to exclude one of the three experimental batches in
the data.  Therefore, I am making three expressionsets, one with all data, and
one the various subsets.

## Only look at wt/cbf5, old and new

In response to some ideas around 20180606, we will extract all the samples which
are wt/cbf5 and not excessively batch-ridden (eg. some of our wt samples)

* all_expt: All samples bundled into one experiment.
* E1_expt: Only the samples from the first experiment.
* E2_expt: Only the smaples from the second experiment.
* E1E2B1_expt: The first experiment and the first batch of the second experiment.
* E1E2B2_expt: The first experiment and the second batch of the second experiment.
* E1E2B2_cbf5wt: Ibid, but only the samples which compare the D95A and wt samples for the CBF5 gene.

```{r gather_all}
all_expt <- sm(create_expt(metadata="sample_sheets/all_samples.xlsx",
                              gene_info=sc_all_annotations,
                              file_column="allfile"))
## Don't forget, I need to change the condition names.
E1_expt <- subset_expt(all_expt, subset="batch=='E1'")
E2_expt <- subset_expt(all_expt, subset="batch!='E1'")
E1E2B2_expt <- subset_expt(all_expt, subset="batch!='E2B1'")
E1E2B1_expt <- subset_expt(all_expt, subset="batch!='E2B2'")

E1E2B2_cbf5wt <- subset_expt(E1E2B2_expt, subset="condition=='WT'|condition=='cbf5_D95A'")
```

# Visualizing raw data

There are lots of methods we have to examine raw data and explore stuff like
batch effects or non-canonical distributions or skewed counts.  hpgltools
provides some functionality to make this process easier.  The graphs shown below
and many more are generated with the wrapper 'graph_metrics()', when invoked it
provides a list of plots including: library sizes, the number of non-zero genes,
density/box plots by sample, distance/correlation heatmaps, standard median
correlation/distance, PCA/TSNE clustering, top-n genes, a legend describing the
colors/symbols used, and some tables describing the data.  Optionally, it can
also perform quantile-quantile plots showing the distribution of each sample
vs. the median of all samples and MA plots of the same.

Caveat: some plots do not work well with gene IDs that are all-0, thus I first
filter the data to remove them.

I also added a neat function 'plot_libsize_prepost()' which shows how many genes
are poorly represented before/after filtering the data.

The plots printed here are all metrics which are useful when considering raw
data.

```{r raw_explore, fig.show="hide"}
all_filt <- sm(normalize_expt(all_expt, filter=TRUE))
all_metrics <- sm(graph_metrics(all_expt))
```

Now print some plots of interest of the raw data.

```{r raw_show}
pp(file="illustrator_input/01_legend.pdf", image=all_metrics$legend)
pp(file="illustrator_input/02_raw_libsize.pdf", image=all_metrics$libsize)

prepost <- plot_libsize_prepost(all_expt)
pp(file="illustrator_input/03_libsize_changed_lowgenes.pdf", image=prepost$lowgene_plot)

pp(file="illustrator_input/04_nonzero_genes.pdf", image=all_metrics$nonzero)
pp(file="illustrator_input/05_raw_boxplot.pdf", image=all_metrics$boxplot)
pp(file="illustrator_input/06_raw_density.pdf", image=all_metrics$density)
pp(file="illustrator_input/07_raw_boxplot.pdf", image=all_metrics$boxplot)
pp(file="illustrator_input/08_topn.pdf", image=all_metrics$topnplot)
```

# Normalize and visualize

Other metrics are more useful when used with data on the log scale and
normalized by number of reads/library and/or by quantile.

```{r normalize, fig.show="hide"}
all_exptm <- normalize_expt(all_expt, transform="log2", convert="cpm", filter=TRUE)
all_metrics <- graph_metrics(all_exptm)
```

The data should now be normalized, lets view some metrics post-facto.

```{r normviz}
pp(file="illustrator_input/09_norm_corheat.pdf", image=all_metrics$corheat)
pp(file="illustrator_input/10_norm_disheat.pdf", image=all_metrics$disheat)
## It appears that just the normalization is sufficient to split the samples completely by type and deeply separate them from the heterologous samples

pp(file="illustrator_input/11_norm_smc.pdf", image=all_metrics$smc)
pp(file="illustrator_input/12_norm_smd.pdf", image=all_metrics$smd)
## The samples are very well behaved, none fall below the red line.

pp(file="illustrator_input/13_norm_pca.pdf", image=all_metrics$pcaplot)
pp(file="illustrator_input/14_norm_tsne.pdf", image=all_metrics$tsneplot)
## The homogeneous wt/mutant are nicely separated, and what is more, the exogeneous samples also split wt/mutant, that might prove to be quite useful.
```

## wt/cbf5 normalized metrics

Above, we split out the wt/cbf5 samples, now lets normalize and make a few plots
of potential interest.  Let us use the same batch method (fsva).

We will do this for both _only_ the old samples and also for the old+new samples.

```{r cbf5_plots, fig.show="hide"}
E1_norm <- normalize_expt(E1_expt, filter=TRUE, norm="quant", convert="cpm", transform="log")

cbf5_norm <- sm(normalize_expt(E1E2B2_cbf5wt, filter=TRUE, norm="quant", convert="cpm",
                               transform="log"))
cbf5_batch <- sm(normalize_expt(E1E2B2_cbf5wt, filter=TRUE, norm="quant", convert="cpm",
                                transform="log", batch="fsva"))

E1_raw_metrics <- sm(graph_metrics(E1_expt))
E1_norm_metrics <- sm(graph_metrics(E1_norm))

cbf5_raw_metrics <- sm(graph_metrics(E1E2B2_cbf5wt))
cbf5_norm_metrics <- sm(graph_metrics(cbf5_norm))
cbf5_batch_metrics <- sm(graph_metrics(cbf5_batch))
```

Now print out some of the interesting plots.

```{r cbf5_show_plots}
pp(file="illustrator_input/32_E1_norm_corheat.pdf", image=E1_norm_metrics$corheat)
pp(file="illustrator_input/33_E1_norm_pca.pdf", image=E1_norm_metrics$pcaplot)
pp(file="illustrator_input/34_E1_norm_disheat.pdf", image=E1_norm_metrics$disheat)
```

## Print some variance partition information

This is an interesting aside which came up last week for some other data.  It
might be good to include the % variance correlated with 'condition' as a column
in the annotation data for the expressionset.  Thus, when we do the differential
expression analysis later, we can look and see if a 'significant' gene has
variance which is actually correlated with condition.

Since writing these analyses, I implemented this idea and so am including it here.

```{r variance_partitions}
vp <- varpart(all_expt, predictor=NULL, factors=c("condition", "batch"))
pp(file="illustrator_input/15_merged_varpart_partition.pdf", image=vp$partition_plot)
## Check out the last two columns!
head(fData(vp$modified_expt))

merged_sva <- sm(normalize_expt(all_expt, transform="log2",
                                filter=TRUE, batch="sva", low_to_zero=TRUE))
vpsva <- varpart(merged_sva, predictor=NULL, factors=c("condition", "batch"))
pp(file="illustrator_input/16_merged_varpart_sva_partition.pdf", image=vpsva$partition_plot)
head(fData(vpsva$modified_expt))
```

# Now look at the data without batch 'E2B1'

```{r nor, fig.show="hide"}
E1E2B2_norm <- sm(normalize_expt(E1E2B2_expt, transform="log2", convert="cpm", norm="quant", filter=TRUE))
E1E2B2_plots <- sm(graph_metrics(E1E2B2_norm))
E1E2B2_batchnorm <- sm(normalize_expt(E1E2B2_expt, transform="log2", convert="cpm", norm="quant", filter=TRUE,
                                      batch="fsva"))
E1E2B2_batchplots <- sm(graph_metrics(E1E2B2_batchnorm))
```

## Print the smaller set of data

```{r E1E2B2_show}
## Before fsva, correlation heatmap without 'r'
pp(file="illustrator_input/17_E1E2B2_corheat.pdf", image=E1E2B2_plots$corheat)
## Before fsva, pca without 'r'
pp(file="illustrator_input/18_E1E2B2_pca.pdf", image=E1E2B2_plots$pcaplot)
## After fsva, correlation heatmap without 'r'
pp(file="illustrator_input/19_E1E2B2_fsva_corheat.pdf", image=E1E2B2_batchplots$corheat)
## After fsva, pca without 'r'
pp(file="illustrator_input/20_E1E2B2_fsva_pca.pdf", image=E1E2B2_batchplots$pcaplot)
```

# Write the expt

```{r fsva_written, fig.show='hide'}
## hmm I was going to silence this line, but looking at the report it seems to be missing some pictures.
all_written <- write_expt(all_expt, excel=paste0("excel/samples_written_all_expt-v", ver, ".xlsx"),
                          filter=TRUE, norm="raw", convert="cpm", batch="fsva", transform="log2")
E1_written <- write_expt(all_expt, excel=paste0("excel/samples_written_E1_expt-v", ver, ".xlsx"),
                         filter=TRUE, norm="raw", convert="cpm", transform="log2")
E1E2B2_written <- write_expt(E1E2B2_expt, excel=paste0("excel/samples_written_E1E2B2_expt-v", ver, ".xlsx"),
                             filter=TRUE, norm="raw", convert="cpm", batch="fsva", transform="log2")

```

```{r saveme, include=FALSE}
if (!isTRUE(get0("skip_load"))) {
  pander::pander(sessionInfo())
  message(paste0("This is hpgltools commit: ", get_git_commit()))
  this_save <- paste0(gsub(pattern="\\.Rmd", replace="", x=rmd_file), "-v", ver, ".rda.xz")
  tmp <- sm(saveme(filename=this_save))
}
```
