---
title: "S. cerevisiae 2017: differential expression, merged sample edition (20180606)."
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
 html_document:
  code_download: true
  code_folding: show
  fig_caption: true
  fig_height: 7
  fig_width: 7
  highlight: default
  keep_md: false
  mode: selfcontained
  number_sections: true
  self_contained: true
  theme: readable
  toc: true
  toc_float:
    collapsed: false
    smooth_scroll: false
---

<style>
  body .main-container {
    max-width: 1600px;
  }
</style>

```{r options, include=FALSE}
if (!isTRUE(get0("skip_load"))) {
  library("hpgltools")
  tt <- devtools::load_all("~/hpgltools")
  knitr::opts_knit$set(progress=TRUE,
                       verbose=TRUE,
                       width=90,
                       echo=TRUE)
  knitr::opts_chunk$set(error=TRUE,
                        fig.width=8,
                        fig.height=8,
                        dpi=96)
  old_options <- options(digits=4,
                         stringsAsFactors=FALSE,
                         knitr.duplicate.label="allow")
  ggplot2::theme_set(ggplot2::theme_bw(base_size=10))
  set.seed(1)
  previous_file <- paste0("02_sample_estimation_merged-v", ver, ".Rmd")
  ver <- "20180606"

  tmp <- sm(loadme(filename=paste0(gsub(pattern="\\.Rmd", replace="", x=previous_file), "-v", ver, ".rda.xz")))
  rmd_file <- paste0("03_differential_expression_merged-v", ver, ".Rmd")
}
```

# Differential Expression: `r ver`

# Filter the raw data

In sample_estimation, I created sc_filt which is precisely what I want.

# Start with batch in the model

I am going to leave these running without silence as I want to make sure they
are running without troubles.

## Set up contrasts

I use the variable 'keepers' to define the numerators/denominators of interest
and give their contrasts names which are appropriate.  I do this because my
all_pairwise() function performs all possible pairwise comparisons in the
specific order of: a:b, a:c, a:d, ... b:c, b:d, ..., c:d, ...
which is not necessarily what is biologically interesting/intuitive.  Thus when
we specifically set the numerators/denominators here, it will make sure that the
result of the contrast is in the chosen orientation.

```{r batch_model}
keepers <- list(
  "D95A_vs_WT" = c("cbf5_D95A", "WT"),
  "upf1d_vs_WT" = c("upf1d", "WT"),
  "double_vs_D95A" = c("cbf5_D95Aupf1d", "cbf5_D95A"),
  "double_vs_upf1d" = c("cbf5_D95Aupf1d", "upf1d"),
  "double_vs_wt" = c("cbf5_D95Aupf1d", "WT"))
```

## Perform search

As I suspect you know, all_pairwise() is the work horse.  It runs
deseq/edger/limma/basic differential expression searches and attempts to use
similar/compatible statistical models for each search (except basic which is
intended as a diagnostic/negative control).

combine_de_tables() does most of the work.  It takes the data from
all_pairwise(), combines them, and writes them out.

I should add some more pictures here.

```{r perform_de}
fsva <- sm(all_pairwise(input=merged_filt, model_batch="fsva", parallel=FALSE))

fsva_write <- combine_de_tables(
  all_pairwise_result=fsva, keepers=keepers,
  excel=paste0("excel/sva_in_model_differential_merged-v", ver, ".xlsx"),
  sig_excel=paste0("excel/sva_in_model_significant-v", ver, ".xlsx"),
  abundant_excel=paste0("excel/sva_in_model_abundance-v", ver, ".xlsx"))
strict_sig_write <- sm(extract_significant_genes(
  fsva_write, lfc=2,
  excel=paste0("excel/sva_in_model_sig2lfc-v", ver, ".xlsx")))

new_colors <- c("#008000", "#4CA64C", "#7FBF7F", "#FF0000", "#FF4C4C", "#FF9999")
new_bars <- significant_barplots(fsva_write, color_list=new_colors)

cbf5_de_plots <- extract_de_plots(fsva_write, type="deseq", table="cbf5_D95A_vs_WT")
pp(file="illustrator_input/21_cbf5_deseq_ma.pdf", image=cbf5_de_plots$ma$plot)
pp(file="illustrator_input/22_cbf5_deseq_vol.pdf", image=cbf5_de_plots$volcano$plot)
pp(file="illustrator_input/23_redgreen_sigbars.pdf", image=new_bars$deseq)
```

## Repeat pairwise searches without the batch 'r'

```{r nor}
merged_nor <- subset_expt(merged_filt, subset="batch!='E2B1'")
fsva_nor <- sm(all_pairwise(input=merged_nor, model_batch="fsva", parallel=FALSE))

fsva_nor_write <- sm(combine_de_tables(
  all_pairwise_result=fsva_nor, keepers=keepers,
  excel=paste0("excel/nor_sva_in_model_differential_merged-v", ver, ".xlsx"),
  abundant_excel=paste0("excel/nor_sva_in_model_abundance-v", ver, ".xlsx"),
  sig_excel=paste0("excel/nor_sva_in_model_significant-v", ver, ".xlsx")))
fsva_nor_sig <- fsva_nor_write[["significant"]]
strict_fsva_nor_write <- sm(extract_significant_genes(
  fsva_nor_write, lfc=2,
  excel=paste0("excel/nor_sva_in_model_sig2lfc-v", ver, ".xlsx")))

new_colors <- c("#008000", "#4CA64C", "#7FBF7F", "#FF0000", "#FF4C4C", "#FF9999")
new_bars <- significant_barplots(fsva_nor_write, color_list=new_colors)

norcbf5_de_plots <- extract_de_plots(fsva_nor_write, type="deseq", table="cbf5_D95A_vs_WT")
pp(file="illustrator_input/24_norcbf5_deseq_ma.pdf", image=norcbf5_de_plots$ma$plot)
pp(file="illustrator_input/25_norcbf5_deseq_vol.pdf", image=norcbf5_de_plots$volcano$plot)
pp(file="illustrator_input/26_norredgreen_sigbars.pdf", image=new_bars$deseq)

pp(file="illustrator_input/27_nor_agreement.pdf", image=fsva_nor$comparison$heat)
```

```{r compare_v1_merged, eval=FALSE, include=FALSE}
v1_env <- new.env()
old_ver <- "20170515"
load(paste0("../scerevisiae_cbf5v1/savefiles/differential_expression-v", old_ver, ".rda.xz"),
     envir=v1_env)
v1_de <- v1_env$batch_write$data$mtc_vs_wtc
##v1_de <- nobatch_write$data$wtc_wtu_vs_mtc_wtu
##rm(list=c("v1_env"))

merged_de <- fsva$deseq$all_tables$wtc_wtu_vs_mtc_wtu
merged_de$logFC <- merged_de$logFC * -1.0
v1v2_merged <- merge(v1_de, merged_de, by="row.names")
v1v2_merged <- v1v2_merged[, c("deseq_logfc", "logFC")]
v1v2_scatter <- plot_linear_scatter(df=v1v2_merged, loess=TRUE)
v1v2_scatter$scatter
v1v2_scatter$correlation
cor.test(v1v2_merged[, 1], v1v2_merged[, 2], method="spearman")
## Well at least the direction is correct...
```

```{r compare_v1_nor, eval=FALSE, include=FALSE}
merged_nor_de <- fsva_nor$deseq$all_tables$wtc_wtu_vs_mtc_wtu
merged_nor_de$logFC <- merged_nor_de$logFC * -1.0
v1v2nor_merged <- merge(v1_de, merged_nor_de, by="row.names")
v1v2nor_merged <- v1v2nor_merged[, c("deseq_logfc", "logFC")]
v1v2nor_scatter <- plot_linear_scatter(df=v1v2nor_merged, loess=TRUE)
v1v2nor_scatter$scatter
v1v2nor_scatter$correlation
cor.test(v1v2_merged[, 1], v1v2_merged[, 2])
## Well at least the direction is correct...
```

# 20180606 Only old data

We want to use our 'only_old' experiment to look at some DE of only experiment 1.

```{r de_expt1}
old_data <- write_expt(only_old, violin=FALSE, batch=FALSE,
                       excel=paste0("excel/only_old_written-v", ver, ".xlsx"))
old_filt <- sm(normalize_expt(only_old, filter=TRUE))
keepers <- list("cbf5_vs_wt" = c("cbf5_D95A", "WT"))
old_de <- sm(all_pairwise(old_filt, model_batch=FALSE, keepers=keepers,
                          combined_excel=paste0("excel/only_old_de-v", ver, ".xlsx"),
                          sig_excel=paste0("excel/only_old_sig-v", ver, ".xlsx")))
```

```{r saveme, include=FALSE}
if (!isTRUE(get0("skip_load"))) {
  pander::pander(sessionInfo())
  message(paste0("This is hpgltools commit: ", get_git_commit()))
  this_save <- paste0(gsub(pattern="\\.Rmd", replace="", x=rmd_file), "-v", ver, ".rda.xz")
  tmp <- sm(saveme(filename=this_save))
}
```
